---
title: "Hydroxyurea Initiation Study — Data Preparation Report"
subtitle: "Ahodwo Facility Patient Records"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 5)

# Anchor working directory to project root regardless of knit location
if (!requireNamespace("rstudioapi", quietly = TRUE) || !rstudioapi::isAvailable()) {
  root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), "../.."))
} else {
  root <- normalizePath(file.path(dirname(rstudioapi::getActiveDocumentContext()$path), "../.."))
}
knitr::opts_knit$set(root.dir = root)
```

```{r packages}
required <- c("dplyr", "ggplot2", "readxl", "tableone", "tidyr", "scales")
missing  <- required[!(required %in% installed.packages()[, "Package"])]
if (length(missing) > 0) install.packages(missing, repos = "https://cloud.r-project.org")

library(dplyr)
library(ggplot2)
library(readxl)
library(tableone)
library(tidyr)
library(scales)

theme_set(
  theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(colour = "grey40"),
          panel.grid.minor = element_blank())
)

BLUE  <- "#2196F3"
RED   <- "#F44336"
GREEN <- "#4CAF50"
```

---

# Overview

This document describes every transformation applied to the raw patient dataset
and provides visual summaries of the final cleaned data. It is intended to
accompany the analysis and serve as an audit trail of all data preparation
decisions made before the prediction model was built.

The goal of the model is to predict which sickle cell disease patients will be
initiated on Hydroxyurea, using their clinical history and longitudinal
laboratory results.

---

# Raw Data — What We Started With

```{r load-raw}
raw <- read_excel("ML/data/data.xlsx", sheet = "Append1")

raw_rows <- nrow(raw)
raw_cols <- ncol(raw)
```

The raw Excel file (`Append1` sheet) contained **`r format(raw_rows, big.mark=",")`
rows** and **`r raw_cols` columns** before any processing.

```{r raw-summary}
data.frame(
  Item  = c("Total rows in raw file", "Total columns", "Sheet name"),
  Value = c(format(raw_rows, big.mark = ","), raw_cols, "Append1")
) |> knitr::kable(col.names = c("", ""))
```

---

# Step-by-Step Manipulations

## Step 1 — Load Data

The file was read directly from Excel using `readxl::read_excel()` targeting the
`Append1` sheet. This preserves original column names (spaces, brackets,
special characters) exactly as they appear in the source file, which are then
cleaned in Step 3.

## Step 2 — Drop Phantom Rows

```{r phantom-rows}
clean <- raw %>% filter(!is.na(`Case ID`) & `Case ID` != "")

phantom_count <- raw_rows - nrow(clean)
```

The bottom **`r format(phantom_count, big.mark=",")`** rows of the file
contained no patient information — only a `Hu Start` flag. These are almost
certainly a formatting artefact from the data export (a secondary summary table
appended to the main records). Including them would skew the outcome balance.

```{r phantom-bar}
data.frame(
  Status = c("Complete records (kept)", "Phantom rows (removed)"),
  Count  = c(nrow(clean), phantom_count)
) %>%
  ggplot(aes(x = Status, y = Count, fill = Status)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = format(Count, big.mark = ",")), vjust = -0.4, fontface = "bold") +
  scale_fill_manual(values = c(GREEN, RED)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Step 2 — Phantom Row Removal",
       subtitle = paste0(format(phantom_count, big.mark = ","),
                         " rows removed — ", format(nrow(clean), big.mark = ","), " remain"),
       x = NULL, y = "Row count")
```

## Step 3 — Standardise Column Names

All 16 columns with messy names (spaces, brackets, slashes, dots) were renamed
to clean snake_case equivalents. Examples:

| Original name | Renamed to |
|---|---|
| `Hu Start(0/1)` | `Hu_Start` |
| `hgb(g/dl)` | `hgb` |
| `rbc *10^12` | `rbc` |
| `Hb Genotype` | `hb_genotype` |
| `form.toxicity` | `toxicity` |
| `history.num_pain_events` | `num_pain_events` |
| `history.experienced_pain` | `experienced_pain` |

Columns that were already clean (e.g. `id_hu_study_number`, `platelet_count`)
were left as-is.

## Step 4 — Drop Columns That Should Never Be Used

```{r drop-cols}
drop_cols <- c(
  "hu_start_date", "patient_full_name", "patient_med_record_num",
  "history.pain_history_notes", "Date Patient Registration",
  "Index", "Column1", "Column2",
  "form.blood_test_results.select_labs",
  "anc (SI Unit) 10^9/L", "anc (Conv. Unit)",
  "arc (SI unit) 10^9/L", "arc (Conv. unit)",
  "wbc (SI Unit) 10^9/L", "hct (SI unit)", "hct(%)"
)

dropped_exist <- drop_cols[drop_cols %in% names(raw)]
```

**`r length(dropped_exist)`** columns were removed:

```{r drop-table}
data.frame(
  Column = c("hu_start_date", "patient_full_name", "patient_med_record_num",
             "history.pain_history_notes", "Date Patient Registration",
             "Index", "Column1", "Column2",
             "form.blood_test_results.select_labs",
             "anc (SI Unit) / (Conv. Unit)",
             "arc (SI unit) / (Conv. unit)",
             "wbc (SI Unit) 10^9/L",
             "hct (SI unit) / hct(%)"),
  Reason = c("Target leakage — directly encodes the outcome",
             "PII — must never enter a model",
             "PII — must never enter a model",
             "98.8% empty; free text, not structured",
             "92.9% empty; too sparse to use",
             "Bare row index, no clinical meaning",
             "Unnamed column, no clinical meaning",
             "Unnamed column, no clinical meaning",
             "Form selector field, no clinical value",
             "Duplicate unit variants of anc (keep primary)",
             "Duplicate unit variants of arc (keep primary)",
             "Duplicate unit variant of wbc (keep primary)",
             "Duplicate unit variants of hct (keep primary)")
) |> knitr::kable()
```

## Step 5 — Filter to Laboratory Visits Only

```{r filter-visits}
after_rename <- clean %>%
  rename(
    Case_ID = `Case ID`, Hu_Start = `Hu Start(0/1)`,
    Lab_Visit_Date = `Lab Visit Date`, `Visit type` = `Visit type`
  )

visit_counts <- table(after_rename$`Visit type`)
```

The dataset contains three visit types. Only **Lab** visits carry blood test
results — the others were excluded.

```{r visit-bar}
as.data.frame(visit_counts) %>%
  rename(Visit_Type = Var1, Count = Freq) %>%
  mutate(Kept = Visit_Type == "Lab") %>%
  ggplot(aes(x = reorder(Visit_Type, -Count), y = Count, fill = Kept)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = format(Count, big.mark = ",")), vjust = -0.4, fontface = "bold") +
  scale_fill_manual(values = c(RED, GREEN)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Step 5 — Visit Type Filter",
       subtitle = "Green = kept (Lab visits only), Red = removed",
       x = "Visit type", y = "Row count")
```

## Step 6 — Remove Duplicates

After filtering to Lab visits, exact duplicate rows were removed using
`duplicated()` across all columns simultaneously. A patient should not appear
twice with identical information on the same date.

## Step 7 — Convert and Standardise Data Types

| Column(s) | Conversion applied |
|---|---|
| `Lab_Visit_Date`, `Clinic_Visit_Date`, `dob` | Excel serial numbers → proper `Date` objects (origin: 30 Dec 1899) |
| `hgb`, `anc`, `arc`, `platelet_count`, `rbc`, `mcv`, `hct`, `wbc` | Text/mixed → numeric; non-convertible entries become `NA` |
| `gender` | Lowercased + trimmed; missing values labelled `"unknown"` → factor |
| `hb_genotype` | Lowercased + trimmed → factor with levels: `ss`, `sc`, `sbo`, `dont_know` |
| `toxicity` | Cast to factor with levels: `none`, `low_anc`, `low_hb`, `low_platelet_count`, `other` |
| `age` | Truncated to whole integer years |

## Step 8 — Handle Missing Lab Values

Missing values in all eight lab columns were replaced with the **column median**
computed from the full dataset.

> **Important note:** This imputation uses all available rows because R's role
> here is data preparation and descriptive reporting only — not model training.
> In Python, imputation will be recomputed on training rows only and then
> applied to the test set separately to prevent any data leakage across the
> train/test split.

## Step 9 — Feature Engineering

Three new sets of features were created:

| Feature | Description |
|---|---|
| `age_group` | `"Paediatric"` (age < 18) or `"Adult"` (age ≥ 18) |
| `days_in_care` | Days elapsed since that patient's first lab visit |
| `hgb_delta` | Difference between current Hgb and the patient's baseline (first visit) Hgb |
| `hgb_prev` | Hgb value from the patient's previous visit |
| `wbc_prev` | WBC value from the patient's previous visit |
| `anc_prev` | ANC value from the patient's previous visit |

For a patient's first visit, `hgb_prev`, `wbc_prev`, and `anc_prev` are `NA`
(no prior visit exists). These were filled with the column median.

---

# Final Dataset — Summary

```{r load-clean}
df <- read.csv("ML/data/hu_clean_data.csv", stringsAsFactors = FALSE)

df$gender      <- as.factor(df$gender)
df$hb_genotype <- as.factor(df$hb_genotype)
df$toxicity    <- as.factor(df$toxicity)
df$age_group   <- as.factor(df$age_group)
df$Hu_Start    <- as.factor(df$Hu_Start)
```

```{r final-summary-table}
data.frame(
  Item  = c("Total rows (lab visits)", "Total columns", "Unique patients",
            "Average visits per patient", "Engineered features added"),
  Value = c(format(nrow(df), big.mark = ","),
            ncol(df),
            format(n_distinct(df$Case_ID), big.mark = ","),
            round(nrow(df) / n_distinct(df$Case_ID), 1),
            6)
) |> knitr::kable(col.names = c("", ""))
```

---

# Visualisations of the Final Clean Data

## Outcome Distribution — Hu_Start

```{r outcome-dist}
df %>%
  count(Hu_Start) %>%
  mutate(pct = n / sum(n),
         label = paste0(format(n, big.mark = ","), "\n(", percent(pct, 1), ")")) %>%
  ggplot(aes(x = Hu_Start, y = n, fill = Hu_Start)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.3, fontface = "bold", lineheight = 1.3) +
  scale_fill_manual(values = c(BLUE, RED)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Outcome Balance — Hu_Start",
       subtitle = "0 = Not on Hydroxyurea  |  1 = Started on Hydroxyurea",
       x = "Hu_Start", y = "Number of visits")
```

The outcome is imbalanced — approximately 78% of visits are from patients on
Hydroxyurea. This will be addressed during modelling with appropriate class
balancing techniques.

## Age Distribution

```{r age-dist}
ggplot(df, aes(x = age)) +
  geom_histogram(binwidth = 2, fill = BLUE, colour = "white") +
  scale_y_continuous(labels = comma) +
  labs(title = "Age Distribution",
       subtitle = "Median age indicates a predominantly paediatric cohort",
       x = "Age (years)", y = "Number of visits") +
  geom_vline(xintercept = median(df$age, na.rm = TRUE),
             linetype = "dashed", colour = RED, linewidth = 0.8) +
  annotate("text", x = median(df$age, na.rm = TRUE) + 1.5,
           y = max(table(cut(df$age, breaks = seq(0, 80, 2)))) * 0.95,
           label = paste0("Median: ", median(df$age, na.rm = TRUE), " yrs"),
           colour = RED, hjust = 0, size = 4)
```

```{r age-group-bar}
df %>%
  count(age_group) %>%
  mutate(pct = n / sum(n),
         label = paste0(format(n, big.mark = ","), " (", percent(pct, 1), ")")) %>%
  ggplot(aes(x = age_group, y = n, fill = age_group)) +
  geom_col(width = 0.4, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.4, fontface = "bold") +
  scale_fill_manual(values = c(BLUE, GREEN)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Age Group — Paediatric vs Adult",
       x = NULL, y = "Number of visits")
```

## Gender Distribution

```{r gender-bar}
df %>%
  count(gender) %>%
  mutate(pct = n / sum(n),
         label = paste0(format(n, big.mark = ","), "\n(", percent(pct, 1), ")")) %>%
  ggplot(aes(x = reorder(gender, -n), y = n, fill = gender)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.3, fontface = "bold", lineheight = 1.3) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Gender Distribution", x = NULL, y = "Number of visits")
```

## Hb Genotype Distribution

```{r genotype-bar}
df %>%
  filter(!is.na(hb_genotype)) %>%
  count(hb_genotype) %>%
  mutate(pct = n / sum(n),
         label = paste0(format(n, big.mark = ","), " (", percent(pct, 1), ")")) %>%
  ggplot(aes(x = reorder(hb_genotype, -n), y = n, fill = hb_genotype)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.4, fontface = "bold") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Haemoglobin Genotype Distribution",
       subtitle = "HbSS dominates — primary candidates for Hydroxyurea",
       x = "Hb Genotype", y = "Number of visits")
```

## Facility Distribution

```{r facility-bar, fig.height=6}
df %>%
  count(Facility_name) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = reorder(Facility_name, n), y = n)) +
  geom_col(fill = BLUE, width = 0.6) +
  geom_text(aes(label = paste0(format(n, big.mark = ","), " (", percent(pct, 1), ")")),
            hjust = -0.05, size = 3.5) +
  coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, 30)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.25))) +
  labs(title = "Visits by Facility", x = NULL, y = "Number of visits")
```

## Laboratory Values — Distributions

```{r lab-distributions, fig.height=10}
lab_long <- df %>%
  select(hgb, anc, arc, platelet_count, wbc, rbc, mcv, hct) %>%
  pivot_longer(everything(), names_to = "Measurement", values_to = "Value")

ggplot(lab_long, aes(x = Value)) +
  geom_histogram(bins = 40, fill = BLUE, colour = "white", alpha = 0.85) +
  facet_wrap(~Measurement, scales = "free", ncol = 2) +
  scale_y_continuous(labels = comma) +
  labs(title = "Laboratory Value Distributions (After Imputation)",
       x = "Value", y = "Count")
```

## Laboratory Values — By Outcome Group

```{r lab-by-outcome, fig.height=10}
df %>%
  select(Hu_Start, hgb, anc, arc, platelet_count, wbc, rbc, mcv, hct) %>%
  pivot_longer(-Hu_Start, names_to = "Measurement", values_to = "Value") %>%
  ggplot(aes(x = Hu_Start, y = Value, fill = Hu_Start)) +
  geom_boxplot(outlier.size = 0.4, alpha = 0.8, show.legend = FALSE) +
  scale_fill_manual(values = c(BLUE, RED)) +
  facet_wrap(~Measurement, scales = "free_y", ncol = 2) +
  labs(title = "Lab Values by Outcome Group",
       subtitle = "0 = Not on Hydroxyurea  |  1 = Started on Hydroxyurea",
       x = "Hu_Start", y = "Value")
```

## Engineered Features

```{r days-in-care}
ggplot(df, aes(x = days_in_care)) +
  geom_histogram(bins = 50, fill = GREEN, colour = "white", alpha = 0.85) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Days in Care Distribution",
       subtitle = "Days elapsed since a patient's first recorded lab visit",
       x = "Days in care", y = "Number of visits")
```

```{r hgb-delta}
ggplot(df, aes(x = hgb_delta)) +
  geom_histogram(bins = 50, fill = BLUE, colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = RED, linewidth = 0.8) +
  scale_y_continuous(labels = comma) +
  labs(title = "Hgb Delta Distribution",
       subtitle = "Change in haemoglobin from each patient's baseline first visit. Negative = falling Hgb.",
       x = "hgb_delta (g/dL)", y = "Number of visits")
```

## Toxicity Categories

```{r toxicity-bar}
df %>%
  filter(!is.na(toxicity)) %>%
  count(toxicity) %>%
  mutate(pct = n / sum(n),
         label = paste0(format(n, big.mark = ","), " (", percent(pct, 1), ")")) %>%
  ggplot(aes(x = reorder(toxicity, -n), y = n, fill = toxicity)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.4, fontface = "bold") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Toxicity / Adverse Event Categories",
       x = NULL, y = "Number of visits")
```

---

# Table 1 — Patient Characteristics at Baseline

One row per patient (earliest lab visit). Stratified by Hu_Start outcome.

```{r table1}
patient_df <- df %>%
  group_by(Case_ID) %>%
  arrange(Lab_Visit_Date) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(Hu_Start = as.factor(Hu_Start))

cont_vars <- c("age", "hgb", "wbc", "anc", "platelet_count")
cat_vars  <- c("gender", "hb_genotype", "Facility_name", "Hu_Start")

table1 <- CreateTableOne(
  vars       = c(cont_vars, cat_vars),
  strata     = "Hu_Start",
  data       = patient_df,
  factorVars = cat_vars,
  addOverall = TRUE
)

print(table1, showAllLevels = TRUE, smd = TRUE)
```

---

# Final Column Inventory

```{r col-inventory}
data.frame(Column = names(df)) |>
  mutate(Type = sapply(df, function(x) class(x)[1])) |>
  knitr::kable()
```

---

*Report generated automatically from `data_cleaning.r` output.*
*Working dataset: `ML/data/hu_clean_data.csv`*
